Approximation as Differentiation

### 2025-07-10 main/language.tex
Do I want to say what a model of List looks like too?

To model Galois slicing semantically for higher-order programs, we define a simple total functional
programming language, extending the simply-typed lambda calculus.

@@

To give the semantics for the language defined in \figrefTwo{syntax}{typing}, we fix a bicartesian closed
category $\Sem$ with finite products $(\times, 1)$, finite coproducts $(+, 0)$ and exponentials
$\internalHom{X}{Y}$, with evaluation morphisms $\eval_{X,Y}$ and currying isomorphisms $\lambda_{X,Y,Z}$,
plus the following additional structure:
\begin{enumerate}
\item a strong monad $(\Lift, \eta, \mu)$ with strength $\mathsf{st}_{X,Y}: X \times \Lift(Y) \to \Lift(X
\times Y)$
\item for each base type $\rho \in \PrimTy$, an object $\sem{\rho}_{\PrimTy}$
\item for each primitive operation $\phi \in \PrimOp^\rho_{\rho_1,\ldots,\rho_n}$, a morphism
$\sem{\phi}_{\Op}: \sem{\rho_1} \times \ldots \times \sem{\rho_n} \to \sem{\rho}$
\item an endofunctor $\List: \Sem \to \Sem$, plus for any object $X$, morphisms $\nil: 1 \to \List(X)$ and
$\cons: X \times \List(X) \to \List(X)$, and for any objects $Y, Z$ and morphisms $f_\nil: Z \to Y$ and
$f_\cons: Z \times X \times Y \to Y$, a morphism $\fold(f_\nil,f_\cons): Z \times \List(X) \to Y$.
\end{enumerate}

@@

\paragraph{Base types and operations}
A \emph{model} of a signature in a category $\cat{C}$ with finite products and a terminal object assigns to
each base type $\rho \in \PrimTy$ an object $\sem{\rho}_{\PrimTy}$ in $\cat{C}$ and to each primitive
operation $\phi \in \PrimOp^\rho_{\rho_1,\ldots,\rho_n}$ a morphism $\sem{\phi}_{\Op}: \sem{\rho_1}_{\PrimTy}
\times \ldots \times \sem{\rho_n}_{\PrimTy} \to \sem{\rho}_{\PrimTy}$.

Given an interpretation of base types and operations in $\Fam(\LatGal)$, the category $\Fam(\MeetSLat \times
\JoinSLat^\op)$ can implement $\Sem$. It is bicartesian closed (\corref{mslat-jslat-bcc}) and can model
$\List(X)$ as an infinite coproduct (\secref{models-of-total-gps:coproducts-and-products}). An interpretation
of base types and operations can be obtained from their interpretation in $\Fam(\LatGal)$ via the embedding
...
\roly{$\Lift$ monad.}

>>>>

To model Galois slicing semantically for higher-order programs, we define a simple total functional
programming language over a signature of base types and primitive operations, extending the simply-typed
lambda calculus.

@@

Models of $\Sigma$ can be given in any category $\cat{C}$ with finite products and a terminal object; such a
model $M$ assigns to each base type $\rho \in \PrimTy$ an object $\sem{\rho}_{M}$ in $\cat{C}$, and to each
primitive operation $\phi \in \PrimOp^\rho_{\rho_1,\ldots,\rho_n}$, a morphism $\sem{\phi}_{M}:
\sem{\rho_1}_{M} \times \ldots \times \sem{\rho_n}_{M} \to \sem{\rho}_{M}$.

To give the semantics for the full language, we fix a bicartesian closed category $\Sem$ plus a model $M$ of
$\Sigma$ in $\Sem$, with the following additional structure:
\begin{enumerate}
\item a strong monad $(\Lift, \eta, \mu)$ with strength $\mathsf{st}_{X,Y}: X \times \Lift(Y) \to \Lift(X
\times Y)$
\item an endofunctor $\List: \Sem \to \Sem$, plus for any object $X$, morphisms $\nil: 1 \to \List(X)$ and
$\cons: X \times \List(X) \to \List(X)$, and for any objects $Y, Z$ and morphisms $f_\nil: Z \to Y$ and
$f_\cons: Z \times X \times Y \to Y$, a morphism $\fold(f_\nil,f_\cons): Z \times \List(X) \to Y$.
\end{enumerate}

@@

\paragraph{Base types and operations}
Given an interpretation of base types and operations in $\Fam(\LatGal)$, the category $\Fam(\MeetSLat \times
\JoinSLat^\op)$ can implement $\Sem$. It is bicartesian closed (\corref{mslat-jslat-bcc}) and can model
$\List(X)$ as an infinite coproduct (\secref{models-of-total-gps:coproducts-and-products}). An interpretation
of base types and operations can be obtained from their interpretation in $\Fam(\LatGal)$ via the embedding
...
\roly{$\Lift$ monad.}
